---
title: "Reproducible Research Assignment W2"
author: "Saurabh Ghadge"
date: "December 5, 2021"
output: html_document
---

#### 1) Reading Data
```{r}
data <- read.csv("repdata_data_activity/activity.csv")
head(data,5)
```

#### 2)Histogram of the total number of steps taken each day
```{r}
library(tidyverse)
data2 <- data %>% group_by(date) %>% summarize(tot_steps = sum(steps,na.rm=TRUE))
ggplot(data=data2)+geom_histogram(mapping = aes(x = tot_steps),na.rm=TRUE)
```

#### 3)Mean and median number of steps taken each day

```{r}
data3 <- data %>% group_by(date) %>% 
        summarise(mean_steps = mean(steps,na.rm = TRUE),
                  med_steps = median(steps,na.rm = TRUE))
data3
```

#### 4)Time series plot of the average number of steps taken
```{r}
data3$date <- as.Date(data3$date)
ggplot(data=data3,mapping = aes(x=date,y=mean_steps))+geom_line(na.rm=TRUE)
```

#### 5)The 5-minute interval that, on average, contains the maximum number of steps
```{r}
data4 <- data %>% group_by(interval) %>% 
        summarise(max_steps = max(steps,na.rm=TRUE))
```

#### 6)Code to describe and show a strategy for imputing missing data
```{r}
any(is.na(data$steps))

```
```{r}
data$steps[which(is.na(data$steps))] <- mean(data$steps,na.rm=TRUE)
any(is.na(data$steps))
```

#### 7)Histogram of the total number of steps taken each day after missing values are imputed
```{r}
data5 <- data %>% group_by(date) %>% summarize(tot_steps = sum(steps))
```

#### 8)Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r warning=FALSE}
#making data weekend and weekdays wise
data$days <- weekdays(as.Date(data$date))
weekend <- grep("Saturday|Sunday",
                data$days,ignore.case = TRUE)
weekend_data <- data[weekend,]
weekend_data$weekday <- "weekend"
weekday_data <- subset(data,data$days != weekend)
weekday_data$weekday <- "weekday"
weekend_day <- rbind(weekend_data,weekday_data)
# for plotting
data7 <- weekend_day %>% group_by(interval,weekday) %>% summarize(mean_steps=mean(steps))
ggplot(data7,mapping = aes(x=interval,y=mean_steps))+geom_point()+facet_grid(weekday~.)
```

